---
title: "ETM640ProjectDraft"
author: "Randolph Anderson"
date: "March 4, 2019"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
The 532d Training Squadron (532 TRS) is a unit in the United States Air Force (USAF).  The 532 TRS
trains the nation's intercontinental ballistic missile (ICBM) operators. The unit also trains
maintenance forces for ICBM and air launched cruise missile (ALCM) systems. 532 TRS instructors
receive professional development (ProD) to increase their knowledge of nuclear enterprise.  
Professional development involves sending one or more nstructors on a trip to a nuclear facility 
in the United States.While on their trip, instructors receive tours and briefings from onsite personnel.
One purpose of providing professional development is to increase effectiveness of instructors. Instructors
can help motivate students by illuminating the importance of what students are learning.  However,
instructors also perform jobs that do not involve actively teaching students.  When in these other
jobs instructors are not in the best position to integrate what they learned into their instruction.

## Problem Definition
Some instructors do not receive professional development until most of their instructing time is
over.  As a result, they never get a chance to share what they have learned with the students.
Since the goal of ProD is, in part, to increase the effectiveness of instructors, the unit may be
able to do a better job assigning instructors to professional development trip in a way that
balances several objectives (goals).

## Objectives/Goals
The squadron has several objectives for the professional development program.  Objectives
include:

1. (Budget) ProD budget should fully utilize remaining travel funds after all mission required
trips are funded; ProD Budget = Total Budget - Mission Required Trips Costs.

2. (Improve Instruction) ProD should help instructors be more effective teaching.

3. (Fair development of instructor core) Each instructor is to be given the opportunity to go on at 
least one ProD trip per year and there should a limit on the max number of trips a person goes on.

4. (Merit):  Instructors merit is considered when assigning instructors to ProD opportunities
that are tagged as being a "glory trip".  A glory trip is does something that is extra special.
The idea is to reward high performing instructors with a special trip.

## Goal Programming Model
Goal programming is an optimization techniques to solve problems with multiple objectives.  
Additionally, Goal programming allows for flexibility in meeting constraints by allowing constraints
to be deviated/broken in order to balance multiple goals.  The model will determine which instructor to 
send on which ProD trip for each year. Goal programming models can be formulated
in many different ways. For example, some models might aim to minimized the sum of deviations from the goal 
targets, while other models, may be set up to minimize the largest deviation (minimax model). This 
project will use a version of the minimax model. A weighted percentage based minimax model is selected for the 
following reasons and assumptions. It is assumed that harm from deviations may increase rapidly such
that it may not be advantage to allow one goal to float to far from target. For example, if objective 3 was 
allowed to deviate it could create harmful perceptions of unfairness. The percentage based minimax
model was selected because harm of deviation was assumed to be more closely related to percentage deviation
then nominal deviation. Weights were applied because squadron leadership has communicated that a loose
prioritization between the goals. Weights will allow the model to capture these priorities into 
the results.

##Decision Variables
There are j instructors and i ProD trips. The model aims to determine which instructors to send on which trip. For each instructor a decision is made to either send or don't send for each trip. To capture all this decisions a discison matrix is defined. $DV[i,j]$ captures the decision for each instructor trip combination. Each element of the matrix is either a 0 or a 1. A 0 indicates that the jth instructor was not assigned to the ith trip. While a 1 indicates that the jth intructor is assigned to the ith trip. For example, DV[1,2]=1 indicates the decision to send instructor 2 to Pro-D trip 1. While DV[12,34] =0 indicates that the 34th instructor is not assigned to 12th trip.

## Objective 
Min: Q, Where Q is less than or equal to the weighted ratio of goal related deviations (e.g Q <= (weight)*Deviation/GoalTarget)

```{r}
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(ggrepel))
suppressPackageStartupMessages(library (magrittr, quietly = TRUE))     #Used for pipes/dplyr
suppressPackageStartupMessages(library (pander, quietly = TRUE))       # for nicely formated tables
suppressPackageStartupMessages(library(dplyr, quietly=TRUE))           # For data structure manipulation
suppressPackageStartupMessages(library(ROI, quietly=TRUE))             # R Optimization Interface package
suppressPackageStartupMessages(library(ROI.plugin.glpk, quietly=TRUE)) # Connection to glpk as solver
suppressPackageStartupMessages(library(ompr, quietly=TRUE))            # Optimization Modeling using R
suppressPackageStartupMessages(library(ompr.roi, quietly=TRUE))        # Connective tissue
suppressPackageStartupMessages(library(tidyr))
```

## Data Input

This is the actual data for instructors in 2018.  It ranks the instructors based on seniority within the squadron.  It designates which instructors were designated as Top Performers as well as which instructors are considered active vs inactive instructors.  The former designation essentially separates those in staff-centric positions and those that primarily fulfill instructor roles.  The 4th quarter hours will not be used in this model but is an additional data point that has potential for usage in future iterations.

```{r Instructor Data}
# reads in the instructor data from a csv file.
RawInstructorData <- as.matrix(read.table(file="InstructorData.csv", row.names=1,sep=","))
colnames(RawInstructorData) <- c("Instructor","DAID", "4QHrs", "Top Performer","Active/Inactive")
RawInstructorData
```

This is the actual information regarding the ProD trips taken in 2018.  It depicts merit-based trips along with the costs and capacities for each trip.  We chose not to use the length of the trip during this iteration, but it could be useful data for future iterations.

```{r Trip Data}
# reads in the instructor data from a csv file.
RawTripData <- as.matrix(read.table(file="ProDTripData.csv", row.names=1,sep=","))
colnames(RawTripData) <- c("Trip Name","MeritBased", "Cost/Instr", "instrSlotsAvailable","NumberOfDays","TotalCost")
RawTripData
```

There are many assumptions to consider within this model.  For instance, we assume that no ProD or mission related trips have been awarded for the current year. This model would be ideally run prior to the beginning of the year vs mid-year.  The total budget and mission related travel costs could be adjusted each year. 

```{r}
TotalTravelBudget <- 188000 #budget for all unit funded travel. This is a hard limit. 
MissionTravelCosts <- 0 #(assumes no mission related trips accomplished prior to model being run. 
ForcastedMissionTravelCosts <- 105000 #current mission related travel forcasted by resource advisor
ProDTravelCosts <- 0  #(assumes no ProD trips accomplished prior to model being run.
TotalRemainingTravelBudget <- 188000-MissionTravelCosts-ProDTravelCosts
ProDTripBudgetTarget <- .9*(TotalTravelBudget-ForcastedMissionTravelCosts) # Target is to use 90% of open funds
NumOfTrips <- length(RawTripData[,1])
NumOfInstructors <- length(RawInstructorData[,1])
TDYCostPerPerson <- as.numeric(RawTripData[,3])
InstructorStatus <- as.numeric(RawInstructorData[,5])
InstructorMerit <-  as.numeric(RawInstructorData[,4])
TripMeritStatus <-  as.numeric(RawTripData[,2])
InstrSlotsAvailable <- as.numeric(RawTripData[,4])
NumOfTripsSlotsAvailable <- sum(InstrSlotsAvailable)

TotalNumOfActiveInstructors<-tabulate(as.numeric(RawInstructorData[,5]),nbins = 1)

#Target for the number of active instructors to send on Trips. 
TotalNumOfProDTripsForActiveInstructorsTarget<-TotalNumOfActiveInstructors*(floor(NumOfTripsSlotsAvailable/NumOfInstructors))+min((NumOfTripsSlotsAvailable%%NumOfInstructors),TotalNumOfActiveInstructors)

#Calc the TargetNumOfInstructorsToSendOnMeritTrips
NumOfMeritTripsSlotsAvailable <- TripMeritStatus %*% InstrSlotsAvailable
NumOfInstructorsWithMeritStatus <- tabulate(as.numeric(RawInstructorData[,4]),nbins = 1)
TargetNumOfInstructorsToSendOnMeritTrips <- min(NumOfInstructorsWithMeritStatus,NumOfMeritTripsSlotsAvailable)
MinNumOfMeritTripsPerTopPerformer <- floor(NumOfMeritTripsSlotsAvailable/NumOfInstructorsWithMeritStatus)
MiniumTripsTarget <- floor(NumOfTripsSlotsAvailable/NumOfInstructors)
MaxTripsTarget <- ceiling(NumOfTripsSlotsAvailable/NumOfInstructors)

Weights <- as.matrix(c(1,1,1,1,1))
```

## Goal Programming Code
```{r MIP Code}

model <- MIPModel() %>%
  
  # VARIABLES
  add_variable (DV[i, j], i=1:NumOfTrips, j=1:NumOfInstructors, type="binary") %>%
  add_variable(BDminus, type = "continuous", lb = 0) %>%
  add_variable(BDpostive, type = "continuous", lb = 0) %>%
  add_variable(IDminus, type = "integer", lb = 0) %>%
  add_variable(IDpostive, type = "integer", lb = 0) %>%
  add_variable(TDminus[j], j=1:NumOfInstructors, type="integer", lb=0) %>%
  add_variable(TDpostive[j], j=1:NumOfInstructors, type="integer", lb=0) %>%
  add_variable(MDminus, type = "integer", lb = 0) %>%
  add_variable(MDpostive, type = "integer", lb = 0) %>%
  
  # OBJECTIVE
  #Minimize the weighted sum of deviations
  set_objective(Weights[1]*(BDminus + BDpostive) + Weights[2]*IDminus +
                  sum_expr(Weights[3]*TDminus[j]+Weights[4]*TDpostive[j], j=1:NumOfInstructors)+Weights[5]*MDminus, "min") %>%
  
  # GOAL CONSTRAINTS
  # 1. ProD Trip Budget Goal
  add_constraint(sum_expr(TDYCostPerPerson[i]*DV[i,j],i=1:NumOfTrips, j=1:NumOfInstructors)
                 + BDminus - BDpostive == ProDTripBudgetTarget)%>%
  
  # 2. Improve instruction Goal
  add_constraint(sum_expr(InstructorStatus[j]*DV[i,j],i=1:NumOfTrips, j=1:NumOfInstructors) + IDminus - IDpostive
                 == TotalNumOfProDTripsForActiveInstructorsTarget)%>%

  # 3. Fairly Develop Instructors: Max Trips
  add_constraint(sum_expr(DV[i,j],i=1:NumOfTrips)-TDpostive[j] <= MaxTripsTarget, j=1:NumOfInstructors)%>%

  # 4. Fairly Develop Instructors: Minimum Trips
  add_constraint(sum_expr(DV[i,j],i=1:NumOfTrips)+TDminus[j] >= MiniumTripsTarget, j=1:NumOfInstructors)%>%

  # 5. Instructor Merit
  add_constraint(sum_expr(InstructorMerit[j]*TripMeritStatus[i]*DV[i,j],i=1:NumOfTrips,j=1:NumOfInstructors) + MDminus -MDpostive  == TargetNumOfInstructorsToSendOnMeritTrips)%>%

  
  # # HARD CONSTRAINTS
  #1. Instructor Merit (if spots are avialble each Top performer will go on a min number of available merit trips)
  add_constraint(sum_expr(InstructorMerit[j]*TripMeritStatus[i]*DV[i,j],i=1:NumOfTrips)
                 >= InstructorMerit[j]*MinNumOfMeritTripsPerTopPerformer,j=1:NumOfInstructors)%>%
  
  # 1. Budget Constraint
  add_constraint(sum_expr(TDYCostPerPerson[i]*DV[i,j],i=1:NumOfTrips, j=1:NumOfInstructors) 
                 <= TotalRemainingTravelBudget)%>%
  
  # 2. Slots Open for Trip
  add_constraint(sum_expr(DV[i,j], j=1:NumOfInstructors) <= InstrSlotsAvailable[i], i=1:NumOfTrips)
  
result <- solve_model(model, with_ROI(solver = "glpk", verbose = TRUE))
```

```{r}

# Active assignments (A) of each jth instructor to ith trips.
A <- result %>% 
  get_solution(DV[i,j]) %>%
  filter(value == 1) %>%  
  select(i, j)

# Text name of the ith trip
TDYNames <- RawTripData[,1]

# Total Num of instructors going on the ith trip
NumAssignedToEachTDY <- tabulate(as.numeric(A[,1]),nbins = NumOfTrips)

# Total Num of top performers going on the ith trip
NumOfTopPerformersAssignedToEachTDY <- matrix(rep(0, each=NumOfTrips), ncol=1)

# Manual counting of all the top performers for each trip.
for(i in 1:nrow(A))
{
  if(A[i,2]*InstructorMerit[A[i,2]] >= 1)
  {
    NumOfTopPerformersAssignedToEachTDY[A[i,1]] = NumOfTopPerformersAssignedToEachTDY[A[i,1]] + 1
  }
}

#TripSummary (TDYName, #Assigned, Capacity, MeritTrip? AssignedOnMerit  Cost/Person)
TripSummaryChart <- cbind(TDYNames,NumAssignedToEachTDY,InstrSlotsAvailable,
                          NumOfTopPerformersAssignedToEachTDY,TripMeritStatus,TDYCostPerPerson) 
colnames(TripSummaryChart)<-list("Names", "Total", "Cap",
                                 "NumOfTPs", "GloryTrip?", "Cost/Per")

# Total Num of instructors going on the ith trip
NumTripsAssignedToEachInstructor <- tabulate(as.numeric(A[,2]),nbins = NumOfInstructors)

# Blank vector to be build up 
NumOfMeritTripsForEachInstructor<- matrix(rep(0, each=NumOfInstructors), ncol=1)

# Of the trips each instructor went on how many trips had merit based classification?
for(t in 1:nrow(A))
{
  if(A[t,1]*TripMeritStatus[A[t,1]] >= 1)
  {
    NumOfMeritTripsForEachInstructor[A[t,2]] = NumOfMeritTripsForEachInstructor[A[t,2]] + 1
  }
}

# The first 3 trips assigned to each instructor
AssignmentByInstructorMatrix <- matrix(rep(0, each=NumOfInstructors*3), ncol=3) #blank matrix to be built

 #Count which trips each instructor went on to a max of 3 trips.
for(k in 1:nrow(A))
{
  # if the first and second slot are filled then use the thrid slot, unless already filled
  if ((AssignmentByInstructorMatrix[A[k,2],3] == 0) && (AssignmentByInstructorMatrix[A[k,2],2] != 0) && (AssignmentByInstructorMatrix[A[k,2],1] != 0))
  {
    AssignmentByInstructorMatrix[A[k,2],3] = A[k,1]
  }
    #if first slot is filled then use the second slot, unless already filled
  if((AssignmentByInstructorMatrix[A[k,2],2] == 0) && (AssignmentByInstructorMatrix[A[k,2],1] != 0))
  {
    AssignmentByInstructorMatrix[A[k,2],2] = A[k,1]
    }
    #if first trip slot isn't filled then record trip, otherwise use next slot
  if(AssignmentByInstructorMatrix[A[k,2],1] == 0)
  {
    AssignmentByInstructorMatrix[A[k,2],1] = A[k,1]
     }
 }

InstructorNumber <- matrix(c(1:NumOfInstructors),ncol=1)
 
# InstructorChart (TotalTrips, MeritTrips, Active?, MeritStatus, Trip1, Trip2, Trip3)
InstructorChart <- cbind(InstructorNumber,NumTripsAssignedToEachInstructor,NumOfMeritTripsForEachInstructor,
                         InstructorMerit,InstructorStatus,AssignmentByInstructorMatrix)
colnames(InstructorChart)<-list("Instructor","TotalTrips", "MeritTrips","MeritStatus","Active?", "Trip1", "Trip2", "Trip3")


# Goal Charts ()
MaxWeightedPercentDeviation <- objective_value(result)
# Goal Charts (Goal,MinusDev, PostiveDev)

# Deviation Results Extacted
BDminusResult <- get_solution(result, BDminus)
BDpostiveResult <-get_solution(result, BDpostive)
IDminusResult <- get_solution(result, IDminus)
IDpostiveResult <- get_solution(result, IDpostive)
MDminusResult <- get_solution(result, MDminus)
MDpostiveResult <- get_solution(result, MDpostive)

# GoalResults
TotalSpent <- ProDTripBudgetTarget - BDminusResult + BDpostiveResult
TotalNumOfProDTripsForActiveInstructors <- TotalNumOfProDTripsForActiveInstructorsTarget - IDminusResult + IDpostiveResult
NumOfTPsSendOnMeritTrips <- TargetNumOfInstructorsToSendOnMeritTrips - MDminusResult + MDpostiveResult

# GoalDataVectors
G1 <- matrix(c(TotalSpent,ProDTripBudgetTarget,BDminusResult,BDpostiveResult),nrow=1)
G2 <- matrix(c(TotalNumOfProDTripsForActiveInstructors,
               TotalNumOfProDTripsForActiveInstructorsTarget,IDminusResult,IDpostiveResult),nrow=1)
G3 <- matrix(c("SeeIntrChart",MiniumTripsTarget,"SeeIntrChart","SeeIntrChart"),nrow=1)
G4 <- matrix(c("SeeIntrChart",MaxTripsTarget,"SeeIntrChart","SeeIntrChart"),nrow=1)
G5 <- matrix(c(NumOfTPsSendOnMeritTrips,TargetNumOfInstructorsToSendOnMeritTrips,MDminusResult,MDpostiveResult),nrow=1)
GoalChart <- rbind(G1,G2,G3,G4,G5)
colnames(GoalChart)<-list("GoalResult","Target","MinusDev","PosDev")
rownames(GoalChart)<-list("Budget","TDYs4ActiveInstructors","MinTripsGoal","MaxTripsGoal", "GloryTrips")

#Weights Data
WeightsByGoal <- t(Weights)
colnames(WeightsByGoal)<-list("Goal1","Goal2","Goal3","Goal4", "Goal5")
rownames(WeightsByGoal)<-list("Weights")

#display the formated results
print(paste0("Objective value= ", objective_value(result)))

pander(WeightsByGoal,
       caption="WeightsSummary")

pander(GoalChart,
       caption="GoalSummary")

pander(TripSummaryChart ,
       caption="TripSummary")

pander(InstructorChart,
       caption="InstructorSummary")
```

All goals were set to equal weights in this iteration. The Goal Summary chart depicts every goal being met or exceeded with the exception of the budget. The target budget for this iteration was arbitrarily set to $\$74,700$.  The model solved for $\$67,954$, deviating by $\$-6,745$. 

The Trip Summary chart show that every trip met the cap exactly with no deviations. All positions for all trips were filled. The instructor summary is interesting.  Only one instructor (#27) was denied a trip.  Placing more wieght on the Minimum Trips Goal (Goal 3) could alter the results to include every instructor. No instructors were allowed a third trip.   All instructors identified as top performers were assigned to trips designated as merit-based trips. 


## 2018 Data Input (Actual Non-Optimized Results)

Below are the results of the actual outcome of the 2018 ProD trips.  Many instructors didn't go on any trips.  Merit trips weren't always matched with top performers. Many in inactive, non-merit status went on three trips.

The optimization model was far more successful in leveling the playing field compared to our current process of selecting instructors for ProD trips. The human element shouldn't be completely discounted, however. For instance, instructor #27 didn't get assigned a trip based on the optimization model results. It would be simple enough to take a trip from an instructor with two and give it to the instructor with zero.   

```{r Actual Instructor Data}
# reads in the instructor data from a csv file.
ActualInstructorData <- as.matrix(read.table(file="2018ActualData.csv", row.names=1,sep=","))
colnames(ActualInstructorData) <- c("TotalTrips", "MeritTrips", "MeritStatus", "Active?", "Trip1", "Trip2", "Trip3")
ActualInstructorData
```
